---
title: RANDOM FOREST IMPUTATION
author:
  - name: Marco R. Steenbergen
    orcid: 0000-0002-5709-5165
    email: steenbergen@ipz.uzh.ch
    affiliations: University of Zurich
date: last-modified
format:
  clean-revealjs:
    revealjs-plugins:
      - tabset
    pdf-separate-fragments: true
bibliography: /Users/prior/Resources/References/master.bib
self-contained: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

## R Packages for this Lecture

Beyond packages you have installed already, please add

- `mice` for missing data imputation with chained equations
- `miceRanger` for implementing multiple imputation
- `missForest` for single RF-based imputations

# Basics of Missing Data {background-color="#40666e"}

## The Bane of the Social Scientist's Existence

- Missing data are a common occurrence in social science data analysis.

- They are common in survey research due to **non-response**.

- In archival work, they occur as well because reporting is not always complete.

## Missing Data Patterns

- Missing data patterns show how often and where missing data occur.

- This is generally a key step for understanding missingness.

## Missingness in the World Indicator Data

::: panel-tabset
## Variables

```{r}
#| echo: true
#| eval: false
library(naniar)
library(rio)
world <- import("Data/world22.xlsx")
gg_miss_var(world, show_pct = TRUE)
```

## Plot1

```{r}
#| message: false
library(naniar)
library(rio)
world <- import("data/world22.xlsx")
gg_miss_var(world, show_pct = TRUE)
```

## Cases

```{r}
#| echo: true
#| eval: false
gg_miss_case(world, order_cases = TRUE)
```

## Plot2

```{r}
gg_miss_case(world, order_cases = TRUE)
```
:::

## Missing Data Mechanisms

- Missing data mechanisms describe how the missing data were generated.

- Generally, three mechanisms are identified [e.g., @rubin1976Inference].

  1.  [**Missing Completely at Random**]{.alert} (MCAR): Missing values arise by chance; no biases are introduced but efficiency suffers.
  2.  [**Missingness at Random**]{.alert} (MAR): Missing values are driven by covariates that may also drive the outcome of interest; biases are introduced and efficiency suffers.
  3.  [**Missingess Not at Random**]{.alert} (MNR): Missing values depend on the values of the outcome of interest; biases are introduced and efficiency suffers.

# Traditional Approaches for Missing Data {background-color="#40666e"}

## Listwise Deletion

- A.k.a. complete case analysis.

- We drop all cases that have any missingness on the variables of interest.

- Unproblematic under MCAR, aside from efficiency loss.

- Bound to induce bias under MAR and MNR.

## Hot Deck Imputation

- The canonical reference is @ford1983Overview.

- Assuming MAR, we substitute the value of the nearest neighbor in terms of a set of covariates.

- This is what we have emulated using kNN.

## Parametric Approaches

- EM algorithm [@dempster1977Maximum].

- Bayesian imputation [@rubin1987Multiple].

- MICE = Multiple Imputation with Chained Equations [@vanbuuren2018Flexible].

- These work quite nicely but

  - Strong assumptions are needed

  - Missingness should not be too great

## MICE Steps

- Imputation stage:
  - We build imputation models appropriate for the type of variable.
  - We run these models $t$ times, each time adding noise proportional to the prediction error from the model.
  - Observed data do not receive noise.
- Analysis stage:
  - We analyze each of the $t$ imputed data sets separately.
  - We average the estimates and adjust their variance by considering within- and between-data set variation.

## MICE with the World Indicators Step 1

::: panel-tabset
## Data

```{r}
#| echo: true
world_imp <- as.data.frame(
  world[, c("DEMINDX22", "HDI21", "EFIDX", "GINI",       # analysis model
            "REGION", "LITERACY", "GDPPERCAP", "GDPGROWTH",
            "POP21", "STABILITY", "GOVEFF", "RULEOFLAW")] # auxiliaries
)
world_imp$log_pop   <- log(world_imp$POP21)
world_imp$log_gdppc <- log(world_imp$GDPPERCAP)
world_imp$POP21 <- world_imp$GDPPERCAP <- NULL
```

## Imputation

```{r}
#| echo: true
#| message: false
library(mice)
set.seed(2026)
imp <- mice(world_imp,
            m       = 20,       # 20 imputed data sets (rule of thumb: >= % missing)
            maxit   = 10,
            printFlag = FALSE)  # defaults: pmm for numeric, polyreg for factors
```

## Diagnostics

```{r}
#| echo: true
#| eval: false
## imputed (red) and observed (blue) distributions should look plausible.
densityplot(imp, ~ DEMINDX22 + HDI21 + EFIDX + GINI)
```

## Plots

```{r}
densityplot(imp, ~ DEMINDX22 + HDI21 + EFIDX + GINI)
```

**Comment:** Observed is blue and imputed = red.
:::

## MICE with the World Indicators Step 2

::: panel-tabset
## Syntax

```{r}
#| echo: true
fit    <- with(imp, lm(DEMINDX22 ~ HDI21 + EFIDX + GINI))
pooled <- pool(fit)
```

## Results

```{r}
#| echo: true
summary(pooled, conf.int = TRUE)
```
:::

# Imputation Using Machine Learning {background-color="#40666e"}

## Why Machine Learning for Missing Data?

- Non-parametric models.

- Capable of learning complex interplay of covariates, as well as nonlinear patterns.

- Tend to work well even when the amount of missing data is large.

## Theory of Missing Forests

- @stekhoven2012MissForest proposed imputation using random forests.

- Training proceeds as follows:

  1.  Initialize by making simple baseline guesses (mean or mode) and by sorting on the amount of missingness $\Rightarrow$ start with least missing data.

  2.  \(a\) Divide the data in to observed and missing entries on a variable. (b) divide all other variables into predictor rows corresponding to the observed and missing parts of the target. (3) Train the forests on the observed data. (4) Predsict and overwrite.

  3.  Stop when matrix differences stabilize over iterations.

## Random Forest Imputation

::: panel-tabset
## Syntax1

```{r}
#| echo: true
#| eval: false
library(missForest)
library(tidyverse)
work <- world |>
  select(-c(COUNTRY, ISO3, REGION, SUBREGION, REGIME))
world_rf_imp <- missForest(work)
print(world_rf_imp$OOBerror)
```

## Results1

```{r}
#| message: false
library(missForest)
library(tidyverse)
work <- world |>
  select(-c(COUNTRY, ISO3, REGION, SUBREGION, REGIME))
world_rf_imp <- missForest(work)
print(world_rf_imp$OOBerror)
```

## Syntax2

```{r}
#| echo: true
world_rf_complete <- world_rf_imp$ximp
```
:::

## The Drawback

- Only one imputed data set is returned: no capture of uncertainty.

- @doove2014Recursive and @shah2014Comparison proposed multiple imputation with RFs.

- Currently, the fastest implementation is `miceRanger`.

## Using `miceRanger`

::: panel-tabset
## Impute

```{r}
#| echo: true
#| message: false
library(miceRanger)
world_imp$REGION <- factor(world_imp$REGION)  # characters must be factors
set.seed(2026)
mr <- miceRanger(world_imp,
                 m       = 20,
                 maxiter = 10,
                 valueSelector       = "meanMatch",  # predictive mean matching on RF predictions
                 meanMatchCandidates = 15,
                 verbose = FALSE)
```

## Checks

```{r}
#| echo: true
#| eval: false
# Ideally, you should run these checks to assess plausibility and
# convergence. For the latter, jaggedness is OK--trending is not.
plotDistributions(mr, vars = c("DEMINDX22", "HDI21", "EFIDX", "GINI"))
plotVarConvergence(mr)   # means/SDs of imputed values across iterations
plotModelError(mr)       # OOB R^2 per variable — how well the forests predict
plotVarImportance(mr)    # which variables drive each imputation model
```

## Model

```{r}
#| echo: true
dat_list <- completeData(mr)
fits <- lapply(dat_list, function(d) lm(DEMINDX22 ~ HDI21 + EFIDX + GINI, data = d))
pooled <- pool(fits)
summary(pooled, conf.int = TRUE)
```
:::

## References
