---
title: MACHINE LEARNING FOR CAUSAL INFERENCE
subtitle: Double Machine Learning and BART
author:
  - name: Marco R. Steenbergen
    orcid: 0000-0002-5709-5165
    email: steenbergen@ipz.uzh.ch
    affiliations: University of Zurich
date: last-modified
format:
  clean-revealjs:
    revealjs-plugins:
      - tabset
    pdf-separate-fragments: true
bibliography: /Users/prior/Resources/References/master.bib
self-contained: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

## R Packages for This Lecture

Beyond packages you have installed already, please add

- `bcf` for Bayesian causal forests.

- `DoubleML` for double machine learning

- `mlr3` a set of efficient machine learning building blocks.

- `ml3learners` for using ranger with double ML

# The Foundations of Causal Inference {background-color="#40666e"}

## The Potential Outcomes Framework

### Rubin Causal Model (RCM)

For each unit $i$, we define two potential outcomes:

- $Y_i(1)$: Outcome if treated ($W_i = 1$)

- $Y_i(0)$: Outcome if control ($W_i = 0$)

The **Individual Treatment Effect (ITE)** is: $$\tau_i = Y_i(1) - Y_i(0)$$

[@rubin1974Estimating; @splawa-neyman1990Application].

## The Fundamental Problem of Causal Inference

- We only observe $Y_i = W_i Y_i(1) + (1 - W_i) Y_i(0)$. The counterfactual is never directly observed [@holland1986Statistics].

- However, under certain assumption it is possible to estimate *average* treatment effects such as:

  - Average treatment effect: $\text{ATE} = \mathbb{E}[Y_i(1) - Y_i(0)]$

  - Average treatment effect of the treated: $\text{ATET} = \mathbb{E}(Y_i(1) - Y_i(0)|W_i = 1)$

## Selection Bias and Unconfoudedness

- If we simply compare observed means ($\mathbb{E}[Y \mid W=1] - \mathbb{E}[Y \mid W=0]$), we get:

  $$\text{Naive Difference} = \underbrace{\text{ATET}}_{\mathbb{E}[Y(1) - Y(0)|W = 1]} + \underbrace{\text{Selection Bias}}_{\mathbb{E}[Y(0) \mid W=1] - \mathbb{E}[Y(0) \mid W=0]}$$

  (confounding).

- Key Identification Assumption: [**Conditional Ignorability**]{.alert}:

  $$(Y(1), Y(0)) \perp W \mid \boldsymbol{x}$$

- We condition on a set of covariates, $\boldsymbol{x}$.

## The Danger of Parametric Controls

In standard empirical workflows, researchers typically write down: $$Y_i = \beta W_i + \boldsymbol{x}_i^\top \boldsymbol{\gamma} + \epsilon_i$$

::: callout-important
## Why this fails in high dimensions

- **Functional Form Misspecification:** Assuming $g(\boldsymbol{x}) = \boldsymbol{x}^\top \boldsymbol{\gamma}$ is linear leaks omitted variable bias into $\hat{\beta}$ if true $g(X)$ is non-linear or interactive.
- **Curse of Dimensionality:** Adding polynomial terms and interactions quickly leads to over-fitting ($P \approx N$).
:::

> **Key takeaway:** We need non-parametric methods to control for $\boldsymbol{X}$ flexibly, without smuggling unexamined linearity assumptions into our causal estimates $\Rightarrow$ machine learning.

# Double Machine Learning {background-color="#40666e"}

## The Historical Anchor

- @robinson1988RootNConsistent introduced the partial linear model:

  $$Y = \beta W + g(\boldsymbol{x}) + U, \quad \mathbb{E}[U \mid W, \boldsymbol{x}] = 0$$

  Where $g(\boldsymbol{x})$ is an **unknown, fully non-parametric function**.

- Robinson's Residualization Algorithm:

  1.  Compute conditional expectations: $m(\boldsymbol{x}) = \mathbb{E}[Y \mid \boldsymbol{x}]$ and $e(\boldsymbol{x}) = \mathbb{E}[W \mid \boldsymbol{X}]$.
  2.  Construct residuals: $\tilde{Y} = Y - m(\boldsymbol{x})$ and $\tilde{W} = W - e(\boldsymbol{x})$.
  3.  Regress $\tilde{Y}$ on $\tilde{W}$ using OLS to obtain a $\sqrt{N}$-consistent estimate of $\beta$.

## Upgrading Robinson: Double Machine Learning (DML)

- @robinson1988RootNConsistent used 1980s kernel smoothers, which collapse when $\boldsymbol{x}$ has more than 3-4 variables.
- Modern Solution [@chernozhukov2018Double]:
  - Replace kernel smoothers with modern ML algorithms (Random Forests, XGBoost, Neural Nets).
- We call this **double** machine learning because we have two predictive tasks:
  - Predicting $Y$ from controls.
  - Predicting $W$ from controls.

## Core Mechanisms

::::: columns
::: {.column width="50%"}
### 1. Cross-Fitting (Sample Splitting)

Prevents over-fitting bias by training ML models for $\hat{m}(\boldsymbol{x})$ and $\hat{e}(\boldsymbol{x})$ on fold $A$, then predicting residuals on fold $B$.
:::

::: {.column width="50%"}
### 2. Neyman-Orthogonality

Ensures that small mistakes in estimating nuisance parameters ($\hat{m}, \hat{e}$) produce zero first-order bias on the estimated parameter $\hat{\beta}$.
:::
:::::

## Illustration in R

::: panel-tabset
## Description

We use an example from @chernozhukov2018Double with simulated, small data.

## Data

```{r}
#| echo: true
#| message: false
# sample splitting with two folds and repeated cross-fitting with n_rep = 2
library(DoubleML)
library(mlr3)
lgr::get_logger("mlr3")$set_threshold("warn")  # silence learner chatter

set.seed(2)
obj_dml_data = make_plr_CCDDHNR2018(n_obs = 500)
```

## Estimation

```{r}
#| echo: true
library(mlr3learners)
learner_l = lrn("regr.ranger", num.trees = 500, min.node.size = 2)
learner_m = lrn("regr.ranger", num.trees = 500, min.node.size = 2)

dml_plr_obj = DoubleMLPLR$new(obj_dml_data,
                              ml_l = learner_l,   # E[Y|X]
                              ml_m = learner_m,   # E[D|X]
                              n_folds = 5,
                              n_rep = 2)

```

## Results 1

```{r}
#| echo: true
dml_plr_obj$fit()
dml_plr_obj$summary()

```

## Results 2

```{r}
#| echo: true
# True effect was 0.5
dml_plr_obj$confint()
```
:::

# Heterogeneous Treatment Effects {background-color="#40666e"}

## Beyond Constant Treatment Effects

- So far, we have assumed there is a single $\beta$ that describes all sub-groups equally well.

- However, treatment effects usually vary.

- How do we build that into machine learning?

## The Meta-Learner Taxonomy

| Learner | Estimator Strategy | Target Objective |
|:-----------------------|:-----------------------|:-----------------------|
| **S-Learner** | Single model $\hat{\mu}(\boldsymbol{x}, W)$ | Predict $Y$ directly; subtract predictions |
| **T-Learner** | Two models: $\hat{\mu}_1(\boldsymbol{x})$ and $\hat{\mu}_0(\boldsymbol{x})$ | Model $W=1$ and $W=0$ independently |
| **X-Learner** | Uses imputed counterfactuals | Designed for extreme treatment imbalance |
| **R-Learner** | Minimizes Robinson's R-Loss | Directly optimizes $(\tilde{Y} - \tilde{W}\tau(\boldsymbol{x}))^2$ |

## The Meta-Learner Taxonomy Cont'd

| Learner | Use Cases | Weaknesses |
|:-----------------------|:-----------------------|:-----------------------|
| **S-Learner** | \(1\) Multi-valued or continuous $W$; (2) RCTs with minimal confounding. | Regularization bias: Trees and regularized models tend to shrink treatment effects toward 0 if main covariate effects dominate $Y$. |
| **T-Learner** | \(1\) Balanced data sets; (2) when you want to ensure $W$ is not ignored. | Inefficient when sample sizes are imbalanced between arms; zero information sharing between arms. |
| **X-Learner** | Severe treatment imbalance. | Computationally heavy. |
| **R-Learner** | Large observational studies requiring **double robustness** (unbiased if *either* the propensity model or outcome model is correctly specified). | Pseudo-outcomes can become volatile and unstable if estimated propensity scores get dangerously close to 0 or 1. |

## Causal Forests

- @athey2019Generalized extend Robinson's residualization trick to **local adaptive neighborhood**.

- Procedure

  1.  Compute global ML predictions $\hat{m}(\boldsymbol{x})$ and $\hat{e}(\boldsymbol{x})$ to obtain residuals $\tilde{Y}$ and $\tilde{X}$.

  2.  Recursively split the feature space into leaf nodes that **maximize heterogeneity** in treatment effect estimates.

  3.  Compute locally weighted OLS inside each leaf:

      $$\hat{tau}(\boldsymbol{x}) = \frac{\sum_{i \in L(x)} \alpha_i(\boldsymbol{x}) \tilde{Y}_i \tilde{W}_i}{\sum_{i \in L(x)} \alpha_i(\boldsymbol{x}) \tilde{W}_i^2}$$

- DR-learning

## Causal Forests Cont'd

> To preserve valid asymptotic inference, trees use separate sub-samples for deciding split locations versus estimating leaf-level treatment effects.

## Regularization-Induced Confounding (RIC)

- In observational studies, when confounding is strong, standard ML trees confuse baseline prognostic variation ($Y(0)$) with treatment heterogeneity ($\tau(\boldsymbol{x})$) [@hahn2020Bayesian].

- For instance,

  - A job-training program has a true positive effect.

  - However, the treatment indicator is greedily used as a shortcut to capture baseline wage differences driven by confounding rather than treatment heterogeneity.

  - As a result the causal effect is shrunk to 0.

## Bayesian Causal Forests

- @hahn2020Bayesian propose Bayesian causal forests (BCF) as a way out off RIC.

- BCF modifies the BART prior:

  $$Y_i = \mu(X_i, \hat{\pi}(\boldsymbol{x}_i)) + W_i \cdot \tau(\boldsymbol{x}_i) + \epsilon_i$$$

- Specifically,

  - **Explicit propensity entry:** Including $\hat{\pi}(\boldsymbol{x})$ directly into $\mu(\cdot)$ absorbs baseline confounding.

  - **Targeted shrinkage:** Applies strong regularization on $\tau(\boldsymbol{x})$ (shrinks to homogeneity) while allowing $\mu(\boldsymbol{x})$ to stay flexible.

## A Hybrid

- **Like a T-Learner / X-Learner:** BCF explicitly separates the treatment groups/effects from the baseline, avoiding the naive blurring of an S-learner.

- **Like an R-Learner / DR-Learner:** BCF directly incorporates an estimated propensity score $\hat{\pi}(X)$ into the model architecture to neutralize confounding.

## Illustration

::: panel-tabset
## DGP

```{r}
#| echo: true
#| eval: false
# data generating process
set.seed(19)

p = 3 #two control variables and W
n = 250

x = matrix(rnorm(n*p), nrow=n)

# create targeted selection
q = -1*(x[,1]>(x[,2])) + 1*(x[,1]<(x[,2]))

# generate treatment variable
pi = pnorm(q)
w = rbinom(n,1,pi)

# tau is the true (homogeneous) treatment effect
tau = (0.5*(x[,3] > -3/4) + 0.25*(x[,3] > 0) + 0.25*(x[,3]>3/4))

# generate the response using q, tau and z
mu = (q + tau*w)

# set the noise level relative to the expected mean function of Y
sigma = diff(range(q + tau*pi))/8

# draw the response variable with additive error
y = mu + sigma*rnorm(n)
```

## Estimation

```{r}
#| echo: true
#| eval: false
#| message: false
library(bcf)
# If you didn't know pi, you would estimate it here
pihat = pnorm(q)

n_burn = 5000
n_sim = 5000

library(bcf)
bcf_fit = bcf(y               = y,
              z               = w,
              x_control       = x,
              x_moderate      = x,
              pihat           = pihat,
              nburn           = n_burn,
              nsim            = n_sim,
              n_chains        = 2,
              update_interval = 100,
              save_tree_directory = './trees')
```
:::

## References
